---
title: "Analysis2025"
output:
  word_document: default
  html_document: default
date: "2025-02-05"
---

```{r}
library(pacman)

p_load(brms, tidyverse, tidybayes, readr, 
       viridis, bayesplot, scales, marginaleffects, cmdstanr)

rhat<-brms::rhat
register_knitr_engine(override = TRUE)
```

All data for each plot

```{r}
prod_data <- read_csv("Hojarasca_CKKEPS.csv") %>%
  rename(Production = "total",
         Site = Plot,
         Plot = CNPlot,
         precip = Rain_F,
         temp = Temp_F,
         vap = Vap_F,
         vpd = VPD_F,
         slope = sl_mean,
         aspect = a_mean) %>%
  mutate(Prod_year = ifelse(Month %in% c(1, 2, 3, 4), 
                            Year - 1, Year),
         Month_t = case_when(Month == 5 ~ 1,
                             Month == 6 ~ 2,
                             Month == 7 ~ 3,
                             Month == 8 ~ 4,
                             Month == 9 ~ 5,
                             Month == 10 ~ 6,
                             Month == 11 ~ 7,
                             Month == 12 ~ 8,
                             Month == 1 ~ 9,
                             Month == 2 ~ 10,
                             Month == 3 ~ 11,
                             Month == 4 ~ 12)) %>%
  arrange(Plot, Prod_year, Month_t) %>%
  group_by(Plot) %>%
  mutate(Time = row_number()) %>%
  ungroup() %>%
  mutate(Plot_year = paste(Plot, Prod_year, sep = "_"),
         Prod_year = factor(Prod_year))

```

Lagged climate data

```{r}

  prod_data <- prod_data %>%
  group_by(Site, Plot) %>%
  mutate(lagged_precip = ifelse(Month %in% c(11, 12, 1, 2, 3, 4, 5), lag(precip), precip),
         lagged_temp = ifelse(Month %in% c(11, 12, 1, 2, 3, 4, 5), lag(temp), temp),
         lagged_vpd = ifelse(Month %in% c(11, 12, 1, 2, 3, 4, 5), lag(vpd), vpd)) %>%
  ungroup()
 
```

Prior experimenting

```{r}
model_min <- brm(bf(Production ~ vpd*catAge + precip*catAge + temp*catAge + s(Month_t, bs = "cc") + 
                 (1|Site) + (1|Plot) + ((vpd + precip + temp)|Prod_year),
                 autocor = ~ ar(time = Time, gr = Plot, p = 1),
                 family = lognormal()),
                 #family = Gamma(link = "log")),
                 #family = gaussian()), 
                 data = prod_data,
                 sample_prior = "only",
                 cores = 4,
                 prior = c(
                   prior(normal(log(0.45), 0.1), class = "Intercept"),
                   prior(normal(0, 0.3), class = "b"),
                   prior(exponential(2), class = "sigma"),
                   prior(exponential(2), class = "sd"),
                   prior(normal(0, 0.3), class = "sds"),
                   prior(normal(0, 0.2), class = "ar"),
                   #prior(gamma(2, 0.1), class = "shape")#,
                   prior(lkj(2), class = "cor")
                   )
                 )

pp_check(model_min, ndraws = 100)
pp_check(model_min, ndraws = 100, type = "stat", stat = "mean")
```


Model specs

```{r}

climate_model <- bf(Production ~ vpd*catAge + precip*catAge + temp*catAge + s(Month_t, bs = "cc") + 
                 (1|Site) + (1|Plot) + ((vpd + precip + temp)|Prod_year),
                 #autocor = ~ ar(time = Time, gr = Plot, p = 1),
                 family = lognormal())

lag_model <- bf(Production ~ lagged_vpd*catAge + lagged_precip*catAge + lagged_temp*catAge + s(Month_t, bs = "cc") + 
                 (1|Site) + (1|Plot) + ((lagged_vpd + lagged_precip + lagged_temp)|Prod_year),
                 autocor = ~ ar(time = Time, gr = Plot, p = 1),
                 family = lognormal())

site_model <- bf(Production ~ arcilla*catAge + pH*catAge + slope*catAge + aspect*catAge + s(Month_t, bs = "cc") + 
                 (1|Site) + (1|Plot) + ((arcilla + pH + slope + aspect)|Prod_year),
                 autocor = ~ ar(time = Time, gr = Plot, p = 1),
                 family = lognormal())

bf_prior <-c(prior(normal(log(0.45), 0.1), class = "Intercept"),
             prior(normal(0, 0.3), class = "b"),
             prior(exponential(2), class = "sigma"),
             prior(exponential(2), class = "sd"),
             prior(normal(0, 0.3), class = "sds"),
             #prior(normal(0, 0.2), class = "ar"),
             prior(lkj(2), class = "cor")
             )

```

Prior check

```{r}

model_prior <- function(model_type, model_name){
  brm(formula = model_type,
      data = prod_data,
      prior = bf_prior,
      init = 0,
      sample_prior = "only",
      iter = 2000,
      warmup = 1000,
      chains = 4,
      cores = 4,
      control = list(adapt_delta = 0.99, max_treedepth = 15),
      file = paste(model_name, "prior", sep = "_"),
      backend = "cmdstanr")}

model_prior(climate_model, "Climate")
model_prior(lag_model, "Lagged")
model_prior(site_model, "Site")

all_priors <- tibble(Type = c("Climate", "Lagged", "Site")) %>%
  mutate(Prior_sampling = map(Type, ~readRDS(paste0(.x,"_prior", ".rds")))) %>%
  mutate(#Prior_overall = map(Prior_sampling, ~pp_check(.x, ndraws = 100, prefix = "ppd")),
         #Prior_mean = map(Prior_sampling, ~pp_check(.x, ndraws = 100, type = "stat", stat = "mean")),
         Prior_scatter = map(Prior_sampling, ~pp_check(.x, ndraws = 100, type = "scatter_avg")))
  
  
walk2(all_priors$Type, all_priors$Prior_overall, 
      ~ggsave(paste0("Prior_check_overall", .x, ".png"),
              .y,
              "png",
              width = 9,
              height = 7,
              units = "in"))

walk2(all_priors$Type, all_priors$Prior_mean, 
      ~ggsave(paste0("Prior_check_mean", .x, ".png"),
              .y,
              "png",
              width = 9,
              height = 7,
              units = "in"))

walk2(all_priors$Type, all_priors$Prior_scatter, 
      ~ggsave(paste0("Prior_check_scatter", .x, ".png"),
              .y,
              "png",
              width = 9,
              height = 7,
              units = "in"))
``` 

Posterior model

```{r}

model_posterior <- function(model_type, model_name){
  brm(formula = model_type,
      data = prod_data,
      prior = bf_prior,
      init = 0,
      iter = 2000,
      warmup = 1000,
      chains = 4,
      cores = 4,
      control = list(adapt_delta = 0.99, max_treedepth = 15),
      file = paste(model_name, "posteriorNOAR", sep = "_"),
      backend = "cmdstanr")}

model_posterior(climate_model, "Climate")
model_posterior(lag_model, "Lagged")
model_posterior(site_model, "Site")
```

Read in posteriors

```{r}
all_posteriors <- tibble(Type = c("Climate", 
                                  "Lagged", 
                                  "Site"
                                  )) %>%
  mutate(posterior_sampling = map(Type, ~readRDS(paste0(.x,"_posterior", ".rds")))) 
```

Posterior predictive checks

```{r}

posterior_checks <- all_posteriors %>%
  mutate(posterior_overall = map(posterior_sampling, ~pp_check(.x, ndraws = 100, prefix = "ppd")),
         posterior_mean = map(posterior_sampling, ~pp_check(.x, ndraws = 100, type = "stat", stat = "mean")),
         posterior_max = map(posterior_sampling, ~pp_check(.x, ndraws = 100, type = "stat", stat = "max")),
         posterior_min = map(posterior_sampling, ~pp_check(.x, ndraws = 100, type = "stat", stat = "min")))
  
  
walk2(all_posteriors$Type, all_posteriors$posterior_overall, 
      ~ggsave(paste0("posterior_check_overall", .x, ".png"),
              .y,
              "png",
              width = 9,
              height = 7,
              units = "in"))

walk2(all_posteriors$Type, all_posteriors$posterior_mean, 
      ~ggsave(paste0("posterior_check_mean", .x, ".png"),
              .y,
              "png",
              width = 9,
              height = 7,
              units = "in"))

walk2(all_posteriors$Type, all_posteriors$posterior_max, 
      ~ggsave(paste0("posterior_check_max", .x, ".png"),
              .y,
              "png",
              width = 9,
              height = 7,
              units = "in"))

walk2(all_posteriors$Type, all_posteriors$posterior_min, 
      ~ggsave(paste0("posterior_check_min", .x, ".png"),
              .y,
              "png",
              width = 9,
              height = 7,
              units = "in"))

```

Model comparison via loo

```{r}

compare <- loo_compare(
  add_criterion((all_posteriors %>%
                   ungroup() %>%
                   filter(Type == "Climate") %>%
                   select(posterior_sampling) %>%
                   pluck("posterior_sampling"))[[1]], "loo", moment_match = TRUE),
  add_criterion((all_posteriors %>%
                   ungroup() %>%
                   filter(Type == "Lagged") %>%
                   select(posterior_sampling) %>%
                   pluck("posterior_sampling"))[[1]], "loo", moment_match = TRUE))

```



lag_model <- bf(Production ~ WindSpeed*catAge + vpd_p*catAge + precip_p*catAge + temp_p*catAge + s(Month_t, bs = "cc") + 
                 (1|Site) + (1|Plot) + (1|Trap) + ((WindSpeed + vpd + precip + temp)|Prod_year),
               autocor = ~ ar(time = Time, gr = Trap, p = 1),
               family = gaussian())

site_model <- bf(Production ~ arcilla*catAge + pH*catAge + slope*catAge + aspect*catAge + s(Month_t, bs = "cc") + 
                 (1|Site) + (1|Plot) + (1|Trap) + ((arcilla + pH + slope + aspect)|Prod_year),
               autocor = ~ ar(time = Time, gr = Trap, p = 1),
               family = gaussian())
               
Climate model

climate_model <- bf(Production ~ vpd*catAge + precip*catAge + temp*catAge + s(Month_t, bs = "cc") + 
                 (1|Site) + (1|Plot) + ((vpd + precip + temp)|Prod_year),
                 family = lognormal())

```{r}
timeseries <- avg_predictions(posterior,
                                 newdata = datagrid(model = posterior,
                                                    Month_t = 1:12,
                                                    Prod_year = unique,
                                                    vpd = mean,
                                                    precip = mean,
                                                    temp = mean,
                                                    arcilla = mean,
                                                    pH = mean,
                                                    slope = mean,
                                                    aspect = mean,
                                                    grid_type = "counterfactual") %>%
                                select(-rowidcf, -Production) %>%
                                distinct(),
                              by = c("Prod_year", "Month_t"))

figure_series<- timeseries %>%
  filter(Prod_year != "2017")%>% 
  ggplot(aes(x = Month_t, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_lineribbon(color = "black", fill = "grey") +
  scale_y_continuous("Biomass production\n") +
  scale_x_continuous("", breaks = 1:12, labels = c("May", "Jun", "Jul", "Aug", 
                                                   "Sep", "Oct", "Nov", "Dec", "Jan", "Feb", "Mar", "Apr")) +
  facet_wrap("Prod_year", nrow = 2) +
  theme_classic() +
  theme(strip.background = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8))

ggsave("Full_biomass.png", device="png", width = 7, height = 7, units = "in", dpi = 600)

figure_year <- timeseries %>%
  filter(Prod_year != "2017", Month_t == 10) %>% 
  ggplot(aes(x = Prod_year, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_bar(stat="identity", color = "black", fill = "grey") +
  geom_errorbar(width=.2) +
  scale_y_continuous("February biomass production\n") +
  scale_x_discrete("") +
  theme_classic() +
  theme(strip.background = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8))

ggsave("Annual_biomass.png", device="png", width = 7, height = 7, units = "in", dpi = 600)

monthseries <- avg_predictions(posterior,
                                 newdata = datagrid(model = posterior,
                                                    Month_t = 1:12,
                                                    Prod_year = unique,
                                                    vpd = mean,
                                                    precip = mean,
                                                    temp = mean,
                                                    grid_type = "counterfactual") %>%
                                 select(-rowidcf, -Production) %>%
                                 distinct() %>%
                                 filter(Prod_year != "2017"),
                              by = "Month_t")
  
figure_month <- monthseries %>% 
  ggplot(aes(x = Month_t, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_lineribbon(color = "black", fill = "grey") +
  scale_y_continuous("Biomass production\n") +
  scale_x_continuous("", breaks = 1:12, labels = c("May", "Jun", "Jul", "Aug", 
                                                   "Sep", "Oct", "Nov", "Dec", "Jan", "Feb", "Mar", "Apr")) +
  theme_classic() +
  theme(strip.background = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8))

ggsave("Monthly_biomass.png", device="png", width = 7, height = 7, units = "in", dpi = 600)
  
```

Calculate slopes for each cat age and year

```{r}

slopes_all <- avg_slopes(posterior, variables = c("vpd", "precip", "temp"),
                         by = c("catAge", "Prod_year"))

figure_slopes <- slopes_all %>% 
  filter(Prod_year !=2017) %>%
  ggplot(aes(x = estimate, y = Prod_year, xmin = conf.low, xmax = conf.high)) +
  geom_pointinterval(aes(color = catAge), position = "dodge") +
  geom_vline(xintercept = 0, linetype = 2) +
  scale_y_discrete("") +
  scale_x_continuous("\nEffect on production") +
  scale_color_viridis("", direction = -1, discrete = TRUE) + 
  facet_grid(. ~ term, scales = "free_x") +
  theme_classic() +
  theme(strip.background = element_blank())

ggsave("Slopes_all.png", device="png", width = 7.5, height = 10, units = "in", dpi = 600)
```

Calculate slopes at mean values for each month or year

```{r}
slopes_rnd <- function(x){
  
  if(x == "Slope"|x == "Age_2021"){
    
    slope_per <-slopes(posterior_mix, variables = x, type = "response", 
                       newdata = datagrid(Prod_month = unique))}else{
                         
    slope_per <-slopes(posterior_mix, variables = x, type = "response", 
                       newdata = datagrid(Prod_year = unique))}}

slopes_final <- tibble(Variable = c("pre", "tmp", "Slope", "Age_2021")) %>%
  group_by(Variable) %>%
  mutate(slopes_going = map(Variable, slopes_rnd))
```

Rate of change in biomass accumulation across spatial gradient per month

```{r}
slope_spatial <- slopes_final %>% 
  ungroup() %>%
  filter(Variable%in%c("Slope", "Age_2021")) %>%
  unnest(slopes_going) %>% 
  mutate(Month = str_split_i(Prod_month, "_", 2),
         Year = str_split_i(Prod_month, "_", 1)) %>%
  mutate(Month = factor(Month, levels = c("May", "June", "July", "August", "September", "October", "November", "December",
                                          "January", "February", "March", "April"))) %>%
  mutate(Variable = ifelse(Variable == "Slope", "Plot slope", "Conglomerate age in 2021")) %>%
  #mutate(Year = as.numeric(Year)) %>%
  ggplot(aes(x = Month, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_pointinterval(aes(color = Year)) +
  geom_hline(yintercept = 0, linetype = 2) +
  scale_y_continuous("Magnitude and direction of effect\n") +
  scale_x_discrete("") +
  scale_color_viridis("", direction = -1, discrete = TRUE) + 
  facet_grid(Variable ~ ., scales = "free_y") +
  theme_classic() +
  theme(strip.background = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8))

#slope_spatial

ggsave("Site_slopes.png", device="png", width = 7, height = 7, units = "in", dpi = 600)

slope_spatial_2 <- slopes_final %>% 
  ungroup() %>%
  filter(Variable%in%c("Slope", "Age_2021")) %>%
  unnest(slopes_going) %>% 
  mutate(Month = str_split_i(Prod_month, "_", 2),
         Year = str_split_i(Prod_month, "_", 1)) %>%
  mutate(Month = case_when(Month == "May" ~ "May",
                           Month == "June" ~ "Jun",
                           Month == "July" ~ "Jul",
                           Month == "August" ~ "Aug",
                           Month == "September" ~ "Sep",
                           Month == "October" ~ "Oct",
                           Month == "November" ~ "Nov",
                           Month == "December" ~ "Dec",
                           Month == "January" ~ "Jan",
                           Month == "February" ~ "Feb",
                           Month == "March" ~ "Mar",
                           Month == "April" ~ "Apr")) %>%
  mutate(Month = factor(Month, levels = c("May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec",
                                          "Jan", "Feb", "Mar", "Apr"))) %>%
  mutate(Variable = ifelse(Variable == "Slope", "Plot slope", "Conglomerate age in 2021")) %>%
  mutate(Year = as.numeric(Year)) %>%
  ggplot(aes(x = Year, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_pointinterval() +
  scale_y_continuous("Magnitude and direction of effect\n") +
  scale_x_continuous("", breaks = seq(2013, 2022, 2)) +
  facet_grid(Variable ~ Month, scales = "free_y") +
  theme_classic() +
  theme(strip.background = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8))

#slope_spatial

ggsave("Site_slopes_2.png", device="png", width = 7, height = 7, units = "in", dpi = 600)


slope_spatial_3 <- slopes_final %>% 
  ungroup() %>%
  filter(Variable%in%c("Slope", "Age_2021")) %>%
  mutate(slopes_posterior = map(slopes_going, ~posterior_draws(.))) %>%
  unnest(slopes_posterior) %>%
  mutate(Month = str_split_i(Prod_month, "_", 2)) %>%
  group_by(Variable, Month) %>%
  median_qi(draw) %>%
  mutate(Month = factor(Month, levels = c("May", "June", "July", "August", "September", "October", "November", "December",
                                          "January", "February", "March", "April"))) %>%
  mutate(Variable = ifelse(Variable == "Slope", "Plot slope", "Conglomerate age in 2021")) %>%
  ggplot(aes(x = Month, y = draw, ymin = .lower, ymax = .upper)) +
  geom_pointinterval() +
  scale_y_continuous("Magnitude and direction of effect\n") +
  scale_x_discrete("") +
  facet_grid(Variable ~ ., scales = "free_y") +
  theme_classic() +
  theme(strip.background = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8))

#slope_spatial

ggsave("Site_slopes_3.png", device="png", width = 7, height = 7, units = "in", dpi = 600)
```


Rate of change in biomass accumulation across monthly gradient per year

```{r}
slope_temporal <- slopes_final %>% 
  ungroup() %>%
  filter(Variable%in%c("tmp", "pre")) %>%
  unnest(slopes_going) %>% 
  mutate(Variable = ifelse(Variable == "tmp", "Temperature", "Precipitation")) %>%
  ggplot(aes(x = Prod_year, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_pointinterval() +
  geom_hline(yintercept = 0, linetype = 2) +
  scale_y_continuous("Magnitude and direction of effect\n") +
  scale_x_discrete("") +
  facet_grid(Variable ~ ., scales = "free_y") +
  theme_classic() +
  theme(strip.background = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8))

#slope_temporal

ggsave("Year_slopes.png", device="png", width = 7, height = 7, units = "in", dpi = 600)

pred_data <- data_mix %>%
  group_by(Prod_year) %>%
  nest() %>%
  mutate(Extremes = map(data, function(x){tibble(tmp = c(rep(mean(x$tmp), 10), 
                                                         seq(min(x$tmp), max(x$tmp), length.out = 10)), 
                                                 pre = c(seq(min(x$pre), max(x$pre), length.out = 10),
                                                         rep(mean(x$pre), 10)))})) %>%
  unnest(Extremes) %>%
  mutate(Slope = mean(data_mix$Slope),
         Age_2021 = mean(data_mix$Age_2021),
         Prod_month = paste(Prod_year, "November", sep = "_"),
         Site = "CN_02",
         Plot = "CN_02_P1",
         Trap =  "CN_02_P1_2")
                                                 
year_preds <- predictions(posterior_mix, newdata = pred_data, variables = c("tmp", "pre"))
```

