---
title: "R code for analyzing CICY decomposition data"
author: "Richard Feldman"
execute: 
  echo: false
format:
  html:
    toc: true
    toc-location: left-body
    toc-title: Contents
    embed-resources: true
    code-overflow: wrap
    self-contained: true
    grid:
      sidebar-width: 200px
      body-width: 1200px
      margin-width: 0px
      gutter-width: 1.5rem
editor: visual
---

```{r}
#| label: load packages
#| message: FALSE
#| warning: FALSE

library(pacman)

p_load(
  brms, tidyverse, tidybayes, readr,
  viridis, bayesplot, scales, marginaleffects, cmdstanr
)

rhat <- brms::rhat
register_knitr_engine(override = TRUE)
```

All data for each plot

```{r}
#| label: load data
#| message: FALSE
#| warning: FALSE

prod_data <- read_csv("Hojarasca_CKKEPS.csv") %>%
  rename(
    Site = CN,
    ID = Plot,
    Plot = ID_Plot
  ) %>%
  mutate(
    Prod_year = ifelse(Month %in% c("Nov", "Dec"),
      Year + 1, Year
    ),
    Month_t = case_when(
      Month == "Nov" ~ 1,
      Month == "Dec" ~ 2,
      Month == "Jan" ~ 3,
      Month == "Feb" ~ 4,
      Month == "Mar" ~ 5,
      Month == "Apr" ~ 6,
      Month == "May" ~ 7,
      Month == "Jun" ~ 8,
      Month == "Jul" ~ 9,
      Month == "Aug" ~ 10,
      Month == "Sept" ~ 11,
      Month == "Oct" ~ 12
    )
  ) %>%
  arrange(Plot, Prod_year, Month_t) %>%
  group_by(Plot) %>%
  mutate(Time = row_number()) %>%
  ungroup() %>%
  mutate(
    Month_year = factor(paste(Prod_year, Month_t, sep = "_")),
    Prod_year = factor(Prod_year)
  )
```

The model: 
1. each month and plot specific decomposition rate depends on that month's temperature
2. decomposition in each plot varies across sites and years
3. the effect of temperature on decomposition varies across years
4. decomposition varies across sites


$$
y_{mpys} = \beta_{0pys} + \beta_{1y} \cdot \text{temp}_{my} + \epsilon_{mpys}, \quad \epsilon_{mpys} \sim \mathcal{N}(0, \sigma^2)\\

\beta_{0pys} = \gamma_{00s} + u_{0ps }+ w_{0y}, \quad u_{0ps} \sim \mathcal{N}(0, \tau_{0}^2), \quad w_{y} \sim \mathcal{N}(0, \omega_{0}^2)\\

\beta_{1y} = \gamma_{10} + v_{1y}, \quad v_{1y} \sim \mathcal{N}(0, \omega_{1}^2)\\

\gamma_{00s} = \delta_{000} + z_{00s}, \quad z_{00s} \sim \mathcal{N}(0, \omega_{s}^2)\\

$$

```{r}
#| label: write model
#| message: FALSE
#| warning: FALSE

priors = c(
    prior(normal(0, 1), class = "Intercept"),
    prior(normal(0, 1), class = "b"),
    prior(exponential(1), class = "sigma"),
    prior(exponential(1), class = "sd"),
    prior(lkj(2), class = "cor")
  )

climate_prior <- function(response){
  formula <-  bf(
  as.formula(paste(response, "~ VPD + Rain + WindSpeed + Temp +
    (1 | Site) + (1 | Plot) + ((VPD + Rain + WindSpeed + Temp) | Prod_year)")),
  autocor = ~ gp(Time, by = Plot),
  family = student
)
  
    brm(
    formula = formula,
    data = prod_data,
    sample_prior = "only",
    prior = priors,
    init = 0,
    iter = 2000,
    warmup = 1000,
    chains = 4,
    cores = 4,
    control = list(adapt_delta = 0.99, max_treedepth = 15),
    file = paste("Climate", response, "prior", sep = "_"),
    backend = "cmdstanr"
  )
}

climate_posterior <- function(response){
  formula <-  bf(
  as.formula(paste(response, "~ VPD + Rain + WindSpeed + Temp +
    (1 | Site) + (1 | Plot) + ((VPD + Rain + WindSpeed + Temp) | Prod_year)")),
  autocor = ~ gp(Time, by = Plot),
  family = student
)
  
    brm(
    formula = formula,
    data = prod_data,
    prior = priors,
    init = 0,
    iter = 2000,
    warmup = 1000,
    chains = 4,
    cores = 4,
    control = list(adapt_delta = 0.99, max_treedepth = 15),
    file = paste("Climate", response, "posterior", sep = "_"),
    backend = "cmdstanr"
  )
}



```

```{r}
#| label: prior predictive check
#| message: FALSE
#| warning: FALSE

prior_model <- climate_prior("Total_Mgha")

pp_check(prior_model, ndraws = 100)
pp_check(prior_model, ndraws = 100, type = "stat", stat = "mean")
```

```{r}
#| label: run posterior model
#| message: FALSE
#| warning: FALSE

climate_total <- climate_posterior("Total_Mgha") 
```

Read in posteriors

```{r}
all_posteriors <- tibble(Type = c(
  "Climate",
  "Lagged",
  "Site"
)) %>%
  mutate(posterior_sampling = map(Type, ~ readRDS(paste0(.x, "_posterior", ".rds"))))
```

Posterior predictive checks

```{r}
posterior_checks <- all_posteriors %>%
  mutate(
    posterior_overall = map(posterior_sampling, ~ pp_check(.x, ndraws = 100, prefix = "ppd")),
    posterior_mean = map(posterior_sampling, ~ pp_check(.x, ndraws = 100, type = "stat", stat = "mean")),
    posterior_max = map(posterior_sampling, ~ pp_check(.x, ndraws = 100, type = "stat", stat = "max")),
    posterior_min = map(posterior_sampling, ~ pp_check(.x, ndraws = 100, type = "stat", stat = "min"))
  )


walk2(
  all_posteriors$Type, all_posteriors$posterior_overall,
  ~ ggsave(paste0("posterior_check_overall", .x, ".png"),
    .y,
    "png",
    width = 9,
    height = 7,
    units = "in"
  )
)

walk2(
  all_posteriors$Type, all_posteriors$posterior_mean,
  ~ ggsave(paste0("posterior_check_mean", .x, ".png"),
    .y,
    "png",
    width = 9,
    height = 7,
    units = "in"
  )
)

walk2(
  all_posteriors$Type, all_posteriors$posterior_max,
  ~ ggsave(paste0("posterior_check_max", .x, ".png"),
    .y,
    "png",
    width = 9,
    height = 7,
    units = "in"
  )
)

walk2(
  all_posteriors$Type, all_posteriors$posterior_min,
  ~ ggsave(paste0("posterior_check_min", .x, ".png"),
    .y,
    "png",
    width = 9,
    height = 7,
    units = "in"
  )
)
```

Model comparison via loo

```{r}
compare <- loo_compare(
  add_criterion((all_posteriors %>%
    ungroup() %>%
    filter(Type == "Climate") %>%
    select(posterior_sampling) %>%
    pluck("posterior_sampling"))[[1]], "loo", moment_match = TRUE),
  add_criterion((all_posteriors %>%
    ungroup() %>%
    filter(Type == "Lagged") %>%
    select(posterior_sampling) %>%
    pluck("posterior_sampling"))[[1]], "loo", moment_match = TRUE)
)
```



lag_model <- bf(Production ~ WindSpeed*catAge + vpd_p*catAge + precip_p*catAge + temp_p*catAge + s(Month_t, bs = "cc") + 
                 (1|Site) + (1|Plot) + (1|Trap) + ((WindSpeed + vpd + precip + temp)|Prod_year),
               autocor = ~ ar(time = Time, gr = Trap, p = 1),
               family = gaussian())

site_model <- bf(Production ~ arcilla*catAge + pH*catAge + slope*catAge + aspect*catAge + s(Month_t, bs = "cc") + 
                 (1|Site) + (1|Plot) + (1|Trap) + ((arcilla + pH + slope + aspect)|Prod_year),
               autocor = ~ ar(time = Time, gr = Trap, p = 1),
               family = gaussian())
               
Climate model

climate_model <- bf(Production ~ vpd*catAge + precip*catAge + temp*catAge + s(Month_t, bs = "cc") + 
                 (1|Site) + (1|Plot) + ((vpd + precip + temp)|Prod_year),
                 family = lognormal())

```{r}
timeseries <- avg_predictions(posterior,
  newdata = datagrid(
    model = posterior,
    Month_t = 1:12,
    Prod_year = unique,
    vpd = mean,
    precip = mean,
    temp = mean,
    arcilla = mean,
    pH = mean,
    slope = mean,
    aspect = mean,
    grid_type = "counterfactual"
  ) %>%
    select(-rowidcf, -Production) %>%
    distinct(),
  by = c("Prod_year", "Month_t")
)

figure_series <- timeseries %>%
  filter(Prod_year != "2017") %>%
  ggplot(aes(x = Month_t, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_lineribbon(color = "black", fill = "grey") +
  scale_y_continuous("Biomass production\n") +
  scale_x_continuous("", breaks = 1:12, labels = c(
    "May", "Jun", "Jul", "Aug",
    "Sep", "Oct", "Nov", "Dec", "Jan", "Feb", "Mar", "Apr"
  )) +
  facet_wrap("Prod_year", nrow = 2) +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8)
  )

ggsave("Full_biomass.png", device = "png", width = 7, height = 7, units = "in", dpi = 600)

figure_year <- timeseries %>%
  filter(Prod_year != "2017", Month_t == 10) %>%
  ggplot(aes(x = Prod_year, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_bar(stat = "identity", color = "black", fill = "grey") +
  geom_errorbar(width = .2) +
  scale_y_continuous("February biomass production\n") +
  scale_x_discrete("") +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8)
  )

ggsave("Annual_biomass.png", device = "png", width = 7, height = 7, units = "in", dpi = 600)

monthseries <- avg_predictions(posterior,
  newdata = datagrid(
    model = posterior,
    Month_t = 1:12,
    Prod_year = unique,
    vpd = mean,
    precip = mean,
    temp = mean,
    grid_type = "counterfactual"
  ) %>%
    select(-rowidcf, -Production) %>%
    distinct() %>%
    filter(Prod_year != "2017"),
  by = "Month_t"
)

figure_month <- monthseries %>%
  ggplot(aes(x = Month_t, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_lineribbon(color = "black", fill = "grey") +
  scale_y_continuous("Biomass production\n") +
  scale_x_continuous("", breaks = 1:12, labels = c(
    "May", "Jun", "Jul", "Aug",
    "Sep", "Oct", "Nov", "Dec", "Jan", "Feb", "Mar", "Apr"
  )) +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8)
  )

ggsave("Monthly_biomass.png", device = "png", width = 7, height = 7, units = "in", dpi = 600)
```

Calculate slopes for each cat age and year

```{r}
slopes_all <- avg_slopes(posterior,
  variables = c("vpd", "precip", "temp"),
  by = c("catAge", "Prod_year")
)

figure_slopes <- slopes_all %>%
  filter(Prod_year != 2017) %>%
  ggplot(aes(x = estimate, y = Prod_year, xmin = conf.low, xmax = conf.high)) +
  geom_pointinterval(aes(color = catAge), position = "dodge") +
  geom_vline(xintercept = 0, linetype = 2) +
  scale_y_discrete("") +
  scale_x_continuous("\nEffect on production") +
  scale_color_viridis("", direction = -1, discrete = TRUE) +
  facet_grid(. ~ term, scales = "free_x") +
  theme_classic() +
  theme(strip.background = element_blank())

ggsave("Slopes_all.png", device = "png", width = 7.5, height = 10, units = "in", dpi = 600)
```

Calculate slopes at mean values for each month or year

```{r}
slopes_rnd <- function(x) {
  if (x == "Slope" | x == "Age_2021") {
    slope_per <- slopes(posterior_mix,
      variables = x, type = "response",
      newdata = datagrid(Prod_month = unique)
    )
  } else {
    slope_per <- slopes(posterior_mix,
      variables = x, type = "response",
      newdata = datagrid(Prod_year = unique)
    )
  }
}

slopes_final <- tibble(Variable = c("pre", "tmp", "Slope", "Age_2021")) %>%
  group_by(Variable) %>%
  mutate(slopes_going = map(Variable, slopes_rnd))
```

Rate of change in biomass accumulation across spatial gradient per month

```{r}
slope_spatial <- slopes_final %>%
  ungroup() %>%
  filter(Variable %in% c("Slope", "Age_2021")) %>%
  unnest(slopes_going) %>%
  mutate(
    Month = str_split_i(Prod_month, "_", 2),
    Year = str_split_i(Prod_month, "_", 1)
  ) %>%
  mutate(Month = factor(Month, levels = c(
    "May", "June", "July", "August", "September", "October", "November", "December",
    "January", "February", "March", "April"
  ))) %>%
  mutate(Variable = ifelse(Variable == "Slope", "Plot slope", "Conglomerate age in 2021")) %>%
  # mutate(Year = as.numeric(Year)) %>%
  ggplot(aes(x = Month, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_pointinterval(aes(color = Year)) +
  geom_hline(yintercept = 0, linetype = 2) +
  scale_y_continuous("Magnitude and direction of effect\n") +
  scale_x_discrete("") +
  scale_color_viridis("", direction = -1, discrete = TRUE) +
  facet_grid(Variable ~ ., scales = "free_y") +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8)
  )

# slope_spatial

ggsave("Site_slopes.png", device = "png", width = 7, height = 7, units = "in", dpi = 600)

slope_spatial_2 <- slopes_final %>%
  ungroup() %>%
  filter(Variable %in% c("Slope", "Age_2021")) %>%
  unnest(slopes_going) %>%
  mutate(
    Month = str_split_i(Prod_month, "_", 2),
    Year = str_split_i(Prod_month, "_", 1)
  ) %>%
  mutate(Month = case_when(
    Month == "May" ~ "May",
    Month == "June" ~ "Jun",
    Month == "July" ~ "Jul",
    Month == "August" ~ "Aug",
    Month == "September" ~ "Sep",
    Month == "October" ~ "Oct",
    Month == "November" ~ "Nov",
    Month == "December" ~ "Dec",
    Month == "January" ~ "Jan",
    Month == "February" ~ "Feb",
    Month == "March" ~ "Mar",
    Month == "April" ~ "Apr"
  )) %>%
  mutate(Month = factor(Month, levels = c(
    "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec",
    "Jan", "Feb", "Mar", "Apr"
  ))) %>%
  mutate(Variable = ifelse(Variable == "Slope", "Plot slope", "Conglomerate age in 2021")) %>%
  mutate(Year = as.numeric(Year)) %>%
  ggplot(aes(x = Year, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_pointinterval() +
  scale_y_continuous("Magnitude and direction of effect\n") +
  scale_x_continuous("", breaks = seq(2013, 2022, 2)) +
  facet_grid(Variable ~ Month, scales = "free_y") +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8)
  )

# slope_spatial

ggsave("Site_slopes_2.png", device = "png", width = 7, height = 7, units = "in", dpi = 600)


slope_spatial_3 <- slopes_final %>%
  ungroup() %>%
  filter(Variable %in% c("Slope", "Age_2021")) %>%
  mutate(slopes_posterior = map(slopes_going, ~ posterior_draws(.))) %>%
  unnest(slopes_posterior) %>%
  mutate(Month = str_split_i(Prod_month, "_", 2)) %>%
  group_by(Variable, Month) %>%
  median_qi(draw) %>%
  mutate(Month = factor(Month, levels = c(
    "May", "June", "July", "August", "September", "October", "November", "December",
    "January", "February", "March", "April"
  ))) %>%
  mutate(Variable = ifelse(Variable == "Slope", "Plot slope", "Conglomerate age in 2021")) %>%
  ggplot(aes(x = Month, y = draw, ymin = .lower, ymax = .upper)) +
  geom_pointinterval() +
  scale_y_continuous("Magnitude and direction of effect\n") +
  scale_x_discrete("") +
  facet_grid(Variable ~ ., scales = "free_y") +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8)
  )

# slope_spatial

ggsave("Site_slopes_3.png", device = "png", width = 7, height = 7, units = "in", dpi = 600)
```


Rate of change in biomass accumulation across monthly gradient per year

```{r}
slope_temporal <- slopes_final %>%
  ungroup() %>%
  filter(Variable %in% c("tmp", "pre")) %>%
  unnest(slopes_going) %>%
  mutate(Variable = ifelse(Variable == "tmp", "Temperature", "Precipitation")) %>%
  ggplot(aes(x = Prod_year, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_pointinterval() +
  geom_hline(yintercept = 0, linetype = 2) +
  scale_y_continuous("Magnitude and direction of effect\n") +
  scale_x_discrete("") +
  facet_grid(Variable ~ ., scales = "free_y") +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8)
  )

# slope_temporal

ggsave("Year_slopes.png", device = "png", width = 7, height = 7, units = "in", dpi = 600)

pred_data <- data_mix %>%
  group_by(Prod_year) %>%
  nest() %>%
  mutate(Extremes = map(data, function(x) {
    tibble(
      tmp = c(
        rep(mean(x$tmp), 10),
        seq(min(x$tmp), max(x$tmp), length.out = 10)
      ),
      pre = c(
        seq(min(x$pre), max(x$pre), length.out = 10),
        rep(mean(x$pre), 10)
      )
    )
  })) %>%
  unnest(Extremes) %>%
  mutate(
    Slope = mean(data_mix$Slope),
    Age_2021 = mean(data_mix$Age_2021),
    Prod_month = paste(Prod_year, "November", sep = "_"),
    Site = "CN_02",
    Plot = "CN_02_P1",
    Trap = "CN_02_P1_2"
  )

year_preds <- predictions(posterior_mix, newdata = pred_data, variables = c("tmp", "pre"))
```
