---
title: "R code for analyzing CICY decomposition data"
author: "Richard Feldman"
execute: 
  echo: false
format:
  html:
    toc: true
    toc-location: left-body
    toc-title: Contents
    embed-resources: true
    code-overflow: wrap
    self-contained: true
    grid:
      sidebar-width: 200px
      body-width: 1200px
      margin-width: 0px
      gutter-width: 1.5rem
editor: visual
---

```{r}
#| label: load packages
#| message: FALSE
#| warning: FALSE

library(pacman)

p_load(
  brms, tidyverse, tidybayes, readr, rlang, ggpubr,
  viridis, bayesplot, scales, marginaleffects, cmdstanr
)

rhat <- brms::rhat
register_knitr_engine(override = TRUE)
```

All data for each plot

```{r}
#| label: load data
#| message: FALSE
#| warning: FALSE

prod_all <- read_csv("Hojarasca_CKKEPS.csv") %>%
  rename(
    Site = CN,
    ID = Plot,
    Plot = ID_Plot
  ) %>%
  mutate(
    Prod_year = ifelse(Month %in% c("Nov", "Dec"),
      Year + 1, Year
    ),
    Site = factor(Site),
    Month_t = case_when(
      Month == "Nov" ~ 1,
      Month == "Dec" ~ 2,
      Month == "Jan" ~ 3,
      Month == "Feb" ~ 4,
      Month == "Mar" ~ 5,
      Month == "Apr" ~ 6,
      Month == "May" ~ 7,
      Month == "Jun" ~ 8,
      Month == "Jul" ~ 9,
      Month == "Aug" ~ 10,
      Month == "Sept" ~ 11,
      Month == "Oct" ~ 12
    )
  ) %>%
  arrange(Plot, Prod_year, Month_t) %>%
  group_by(Plot) %>%
  mutate(Time = row_number()) %>%
  ungroup() %>%
  mutate(
    Month_year = factor(paste(Prod_year, Month_t, sep = "_")),
    Plot_year = paste(Plot, Prod_year, sep = "_")
  )
```

The model: 
1. each month and plot specific decomposition rate is modeled as a function of that month's temperature
2. residual variation in the decomposition rate includes both temporally correlated structure and independent noise
3. the temporally correlated residual component is modeled as a Gaussian Process over time within each plot
4. decomposition rate in each plot varies across sites and years
5. the effect of temperature on decomposition varies across years
4. decomposition rate varies across sites


$$
y_{mpys} = \beta_{0pys} + \beta_{1y} \cdot \text{temp}_{my} +\alpha_{mpys} +\epsilon_{mpys}\\ \quad \epsilon_{mpys} \sim \mathcal{N}(0, \sigma^2)\\\quad \alpha_{mpys} \sim GP(0, K(t,t'))\\K(t,t')=\eta^2exp(-\frac{(time_i-time_j)^2}{2rho^2})\\

\beta_{0pys} = \gamma_{00s} + u_{0ps }+ w_{0y}\\ \quad u_{0ps} \sim \mathcal{N}(0, \tau_{0}^2)\\ \quad w_{y} \sim \mathcal{N}(0, \omega_{0}^2)\\

\beta_{1y} = \gamma_{10} + v_{1y}\\ \quad v_{1y} \sim \mathcal{N}(0, \omega_{1}^2)\\

\gamma_{00s} = \delta_{000} + z_{00s}\\ \quad z_{00s} \sim \mathcal{N}(0, \omega_{s}^2)\\

$$
Method 1. Remove 2017 and 2018 with the missing covariates

```{r}
#| label: write model
#| message: FALSE
#| warning: FALSE

prod_data <- prod_all %>%
  filter(!Prod_year %in% c(2017, 2018)) %>%
  mutate(Prod_year = factor(Prod_year))

priors <- c(
  prior(normal(0, 1), class = "Intercept"),
  prior(normal(0, 1), class = "b"),
  prior(exponential(1), class = "sigma"),
  prior(exponential(1), class = "sd"),
  prior(lkj(2), class = "cor")
)

climate_prior <- function(response) {
  formula <- bf(
    as.formula(paste(response, "~ VPD + Rain + WindSpeed + Temp +
  (1 | Site) + (1 | Plot) + ((VPD + Rain + WindSpeed + Temp) | Prod_year)")),
    autocor = ~ gp(Month_t, by = Plot_year),
    family = student
  )


  brm(
    formula = formula,
    data = prod_data,
    sample_prior = "only",
    prior = priors,
    init = 0,
    iter = 2000,
    warmup = 1000,
    chains = 4,
    cores = 4,
    control = list(adapt_delta = 0.99, max_treedepth = 15),
    file = paste("Climate", response, "prior", sep = "_"),
    backend = "cmdstanr"
  )
}

climate_posterior <- function(response) {
  formula <- bf(
    as.formula(paste(response, "~ VPD + Rain + WindSpeed + Temp +
    (1 | Site) + (1 | Plot) + ((VPD + Rain + WindSpeed + Temp) | Prod_year)")),
    autocor = ~ gp(Month_t, by = Plot_year),
    family = student
  )

  brm(
    formula = formula,
    data = prod_data,
    prior = priors,
    init = 0,
    iter = 2000,
    warmup = 1000,
    chains = 4,
    cores = 4,
    control = list(adapt_delta = 0.99, max_treedepth = 15),
    file = paste("Climate", response, "posterior", sep = "_"),
    backend = "cmdstanr"
  )
}
```

Method 2. impute the missing covariates

```{r}
#| label: missing imputation model
#| message: FALSE
#| warning: FALSE

set_priors <- function(response) {
  c(
    # Priors for imputed covariates
    prior(normal(0, 5), class = "b", coef = "miRain", resp = response),
    prior(normal(0, 5), class = "b", coef = "miTemp", resp = response),
    prior(normal(0, 5), class = "b", coef = "miVPD", resp = response),
    prior(normal(0, 5), class = "b", coef = "miWindSpeed", resp = response),

    # Optional: Priors for intercepts and other fixed effects
    prior(student_t(3, 0, 10), class = "Intercept", resp = response),
    prior(student_t(3, 0, 10), class = "sigma", resp = response),

    # Priors for random effect standard deviations
    prior(exponential(1), class = "sd", group = "Plot", resp = response),
    prior(exponential(1), class = "sd", group = "Site", resp = response),
    prior(exponential(1), class = "sd", group = "Prod_year", resp = response)
  )
}

climate_prior <- function(response) {
  priors <- c(
    # Priors for imputed covariates
    prior(normal(0, 5), class = "b", coef = "miRain", resp = response),
    prior(normal(0, 5), class = "b", coef = "miTemp", resp = response),
    prior(normal(0, 5), class = "b", coef = "miVPD", resp = response),
    prior(normal(0, 5), class = "b", coef = "miWindSpeed", resp = response),

    # Optional: Priors for intercepts and other fixed effects
    prior(student_t(3, 0, 10), class = "Intercept", resp = response),
    prior(student_t(3, 0, 10), class = "sigma", resp = response),

    # Priors for random effect standard deviations
    prior(exponential(1), class = "sd", group = "Plot", resp = response),
    prior(exponential(1), class = "sd", group = "Site", resp = response),
    prior(exponential(1), class = "sd", group = "Prod_year", resp = response)
  )
  formula <- bf(
    as.formula(paste(response, "|mi() ~ mi(VPD) + mi(Rain) + mi(WindSpeed) + mi(Temp) +
    (1 | Site) + (1 | Plot) + ((mi(VPD) + mi(Rain) + mi(WindSpeed) + mi(Temp)) | Prod_year)")),
    autocor = ~ gp(Time, by = Plot)
  )

  brm(
    formula = formula +
      bf(VPD | mi() ~ 1) +
      bf(Temp | mi() ~ 1) +
      bf(Rain | mi() ~ 1) +
      bf(WindSpeed | mi() ~ 1) +
      set_rescor(FALSE),
    family = student,
    data = prod_data,
    sample_prior = "only",
    prior = priors,
    init = 0,
    iter = 2000,
    warmup = 1000,
    chains = 4,
    cores = 4,
    control = list(adapt_delta = 0.99, max_treedepth = 15),
    file = paste("Climate_imputed", response, "prior", sep = "_"),
    backend = "cmdstanr"
  )
}

climate_posterior <- function(response) {
  formula <- bf(
    as.formula(paste(response, "~ VPD + Rain + WindSpeed + Temp +
    (1 | Site) + (1 | Plot) + ((VPD + Rain + WindSpeed + Temp) | Prod_year)")),
    autocor = ~ gp(Time, by = Plot),
    family = student
  )

  brm(
    formula = formula,
    data = prod_data,
    prior = priors,
    init = 0,
    iter = 2000,
    warmup = 1000,
    chains = 4,
    cores = 4,
    control = list(adapt_delta = 0.99, max_treedepth = 15),
    file = paste("Climate", response, "posterior", sep = "_"),
    backend = "cmdstanr"
  )
}
```


```{r}
#| label: set-up models
#| message: FALSE
#| warning: FALSE

climate_setup <- tibble(Measure = c("HG_Mgha", "HF_Mgha", "RG_Mgha", "RF_Mgha", "ER_Mgha", "Total_Mgha"))
```

```{r}
#| label: run prior models
#| message: FALSE
#| warning: FALSE

climate_prior("Total_Mgha")
```

```{r}
#| label: prior predictive check
#| message: FALSE
#| warning: FALSE

all_priors <- climate_setup %>%
  filter(Measure == "Total_Mgha") %>%
  mutate(Prior_sampling = map(Measure, ~ readRDS(paste0("Climate_", .x, "_prior", ".rds")))) %>%
  mutate(
    Prior_overall = map(Prior_sampling, ~ pp_check(.x, ndraws = 100, prefix = "ppd")),
    Prior_mean = map(Prior_sampling, ~ pp_check(.x, ndraws = 100, Measure = "stat", stat = "mean")),
    Prior_scatter = map(Prior_sampling, ~ pp_check(.x, ndraws = 100, Measure = "scatter_avg"))
  )


walk2(
  all_priors$Measure, all_priors$Prior_overall,
  ~ ggsave(paste0("Prior_climate_overall", .x, ".png"),
    .y,
    "png",
    width = 9,
    height = 7,
    units = "in"
  )
)

walk2(
  all_priors$Measure, all_priors$Prior_mean,
  ~ ggsave(paste0("Prior_climate_mean", .x, ".png"),
    .y,
    "png",
    width = 9,
    height = 7,
    units = "in"
  )
)

walk2(
  all_priors$Measure, all_priors$Prior_scatter,
  ~ ggsave(paste0("Prior_climate_scatter", .x, ".png"),
    .y,
    "png",
    width = 9,
    height = 7,
    units = "in"
  )
)
```

```{r}
#| label: run posterior model
#| message: FALSE
#| warning: FALSE

climate_posterior("Total_Mgha")
```

```{r}
#| label: run posterior predictive checks
#| message: FALSE
#| warning: FALSE

all_posteriors <- climate_setup %>%
  filter(Measure == "Total_Mgha") %>%
  mutate(Posterior_sampling = map(Measure, ~ readRDS(paste0("Climate_", .x, "_posterior", ".rds")))) %>%
  mutate(
    Posterior_overall = map(Posterior_sampling, ~ pp_check(.x, ndraws = 100, prefix = "ppd")),
    Posterior_mean = map(Posterior_sampling, ~ pp_check(.x, ndraws = 100, type = "stat", stat = "mean")),
    Posterior_max = map(Posterior_sampling, ~ pp_check(.x, ndraws = 100, type = "stat", stat = "max")),
    Posterior_min = map(Posterior_sampling, ~ pp_check(.x, ndraws = 100, type = "stat", stat = "min"))
  )


walk2(
  all_posteriors$Measure, all_posteriors$Posterior_overall,
  ~ ggsave(paste0("Posterior_climate_overall", .x, ".png"),
    .y,
    "png",
    width = 9,
    height = 7,
    units = "in"
  )
)

walk2(
  all_posteriors$Measure, all_posteriors$Posterior_mean,
  ~ ggsave(paste0("Posterior_climate_mean", .x, ".png"),
    .y,
    "png",
    width = 9,
    height = 7,
    units = "in"
  )
)

walk2(
  all_posteriors$Measure, all_posteriors$Posterior_max,
  ~ ggsave(paste0("Posterior_climate_max", .x, ".png"),
    .y,
    "png",
    width = 9,
    height = 7,
    units = "in"
  )
)

walk2(
  all_posteriors$Measure, all_posteriors$Posterior_min,
  ~ ggsave(paste0("Posterior_climate_min", .x, ".png"),
    .y,
    "png",
    width = 9,
    height = 7,
    units = "in"
  )
)
```


```{r}
#| label: Set-up predicted decomposition rates and spatial and temporal variation in slopes
#| message: FALSE
#| warning: FALSE

climate_vars <- c("VPD", "Temp", "Rain", "WindSpeed")

climate_means <- prod_data %>%
  select(climate_vars) %>%
  summarize(
    VPD = mean(VPD, na.rm = T),
    Temp = mean(Temp, na.rm = T),
    Rain = mean(Rain, na.rm = T),
    WindSpeed = mean(WindSpeed, na.rm = T)
  )

# Predictions

climate_predictions <- function(results, variable) {
  pred_var <- prod_data %>%
    select(variable)

  pred_data <- expand_grid(
    seq(min(pred_var, na.rm = T),
      max(pred_var, na.rm = T),
      length.out = 100
    ),
    Plot = unique(prod_data$Plot),
    Prod_year = unique(prod_data$Prod_year)
  )

  names(pred_data)[1] <- names(pred_var)

  other_means <- prod_data %>%
    select(Site, Plot) %>%
    distinct() %>%
    bind_cols(climate_means) %>%
    select(-variable)

  pred_final <- full_join(pred_data, other_means, by = "Plot")

  avg_predictions(results,
    newdata = pred_final, by = variable
  )
}

# Figure - predictions

figure_predictions <- function(variable, predictions) {
  var_sym <- sym(variable)

  predictions %>%
    ggplot(aes(x = !!var_sym, y = estimate, ymin = conf.low, ymax = conf.high)) +
    geom_lineribbon(color = "black", fill = "grey") +
    scale_y_continuous("Decomposition rate\n") +
    scale_x_continuous(name = paste0("\n", variable)) +
    theme_classic() +
    theme(axis.title.y = element_blank())
}


# Slopes

climate_slopes <- function(results, variable, type) {
  avg_slopes(results,
    variables = variable, by = type
  )
}

figure_slopes <- function(variable, type, slopies) {
  type_sym <- sym(type)

  slopies %>%
    ggplot(aes(x = estimate, y = !!type_sym, xmin = conf.low, xmax = conf.high)) +
    geom_pointinterval(color = "burlywood4", point_size = 3) +
    scale_y_discrete("") +
    scale_x_continuous(name = paste0("\n", variable)) +
    theme_classic()
}
```

```{r}
#| label: Figures of decomposition rates and slopes
#| message: FALSE
#| warning: FALSE

all_predictions <- expand_grid(all_posteriors %>%
  select(Measure, Posterior_sampling), climate_vars) %>%
  mutate(preds = map2(Posterior_sampling, climate_vars, climate_predictions)) %>%
  mutate(figures = map2(climate_vars, preds, figure_predictions))

annotate_figure(
  ggarrange(plotlist = all_predictions$figures),
  left = "Decomposition rate\n"
)

all_slopes <- expand_grid(all_posteriors %>%
  select(Measure, Posterior_sampling), climate_vars, Type = c("Site", "Plot", "Prod_year")) %>%
  mutate(slopes = pmap(list(Posterior_sampling, climate_vars, Type), climate_slopes)) %>%
  mutate(figures = pmap(list(climate_vars, Type, slopes), figure_slopes))

annotate_figure(
  ggarrange(plotlist = all_slopes %>%
    filter(Type == "Site") %>%
    pull(figures)),
  left = "Site\n",
  bottom = "\nEffect on decomposition rate"
)

annotate_figure(
  ggarrange(plotlist = all_slopes %>%
    filter(Type == "Prod_year") %>%
    pull(figures)),
  left = "Year (Nov - Oct)\n",
  bottom = "\nEffect on decomposition rate"
)
```
